Towards Comparable Active Learning

### 2024-06-03 doc/al_benchmark_paper.tex
paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Description}
Given two spaces $\X, \Y$, $n=l+u$ data points with $l \in \mathbb{N}$ labeled examples $\mathcal{L} = \{(x_1, y_1),\ldots, (x_l,y_l)\}$, $u \in \mathbb{N}$ unlabeled examples $\mathcal{U} = \{x_{l+1},\ldots,x_{n}\}$, a model $\hat y: \X \to \Y$, a budget $\mathbb{N} \ni b \le u$ and an annotator $A: \mathcal{X} \to \mathcal{Y}$ that can label $x$ (we consider only hard labels in the one-hot format). 
We call $x \in \mathcal{X}$, $y \in \mathcal{Y}$ predictors and labels respectively where $(x,y)$ are drawn from an unknown distribution $\rho$. 
Find an acquisition function $\Omega: \U^{(i)},\LL^{(i)} \mapsto x^{(i)} \in \U^{(i)}$ that iteratively selects the next unlabeled point $x^{(i)}$ for labeling
\begin{align*}
	\LL^{(i+1)} &\gets \LL^{(i)} \cup \{\left(x^{(i)}, A(x^{(i)})\right)\} \\
	\U^{(i+1)} &\gets \U^{(i)} \setminus x^{(i)} %, i \in 1{:}B
\end{align*}
with $\U^{(0)} = \operatorname{seed}(\U, s)$ and $\LL^{(0)} = \{\left(\U^{(0)}, A(\U^{(0)})\right)\}$, where $\operatorname{seed}(\U, s)$ selects $s$ points per class for the initial labeled set. \\
So that the average expected loss $\ell: \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$ of a machine learning algorithm fitting $\hat y^{(i)}$ on the respective labeled set $\LL^{(i)}$ is minimal: 
$$\min \quad \frac{1}{B} \sum\limits_{i=0}^B \mathbb{E}_{(x,y) \sim \rho} \ell(y, \hat y^{(i)})$$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why we need 50 restarts}\label{sec:restarts}
To evaluate how many restarts are necessary to obtain conclusive and reproducible results in an AL experiment, we computed 100 runs of our top-performing algorithm on one dataset.
Our best algorithm is margin sampling and we chose the Splice dataset for its average size and complexity. \\
This allows us firstly, to obtain a very strong estimation of the "true" average performance of margin sampling on this dataset and secondly, to draw subsets from this pool of 100 runs.
Setting the size of our draws to $\alpha$ and sampling uniformly, we can approximate a cross-validation process with $\alpha$ restarts.
Each of these draws can be interpreted as a \textbf{reported result in AL literature} where the authors employed $\alpha$ restarts.
%We provide evidence in Fig. \ref{fig:restarts} that previous works might not have evaluated their experiments with a sufficient number of restarts.
Figure \ref{fig:restarts} shows the "true" mean performance of margin sampling (green) in relation to random sampling (black) and the oracle performance (red).
We display 5 random draws of size $\alpha$ in blue.
We can observe that even for a relatively high number of restarts the variance between the samples is extremely high, resulting in some performance curves being worse that random and some being significantly better.
When setting $\alpha = 50$ we observe all samples to converge close to the true mean performance. 
\begin{figure}
        \hspace{-20mm}
	\includegraphics[width=1.24\linewidth]{img/ablation_restarts}
	\caption{Random draws from a pool of 100 runs for margin sampling on the Splice dataset with different numbers of repetitions ($\alpha=\{5,10,50\}$). Green curves are the mean performance of all 100 runs, while the samples are blue. Even with 5 or 10 repetitions, we can observe that single draws for margin sampling display below-random performance (black), while the true mean should be above random.}
	\label{fig:restarts}
\end{figure}
In addition to this motivating example, we carried out our main evaluation (Tab. \ref{tab:results}) multiple times by uniformly sampling 3 random from our 50 available runs and comparing the results.
We found significant differences in the performance of acquisition functions on individual datasets, as well as permutations in the final ranking.
This partly explains the ongoing difficulties in reproducing results for AL experiments and benchmarks.
This details can be found in App. \ref{app:rank_difference}.
For this benchmark we employ 50 restarts of every experiment.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation Protocol}\label{sec:evaluation}
Following \cite{zhou2021towards}, the quality of an AL algorithm is evaluated by an ``anytime protocol" that incorporates classification performance at every iteration, as opposed to evaluating final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{AUC}(\D_\test, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(\D_\test, \hat y^{(i)})
\end{equation}
The AUC incorporates performance in early stages (low budget) as well as capabilities to push the classifier in later stages (high budget).
AL algorithms have to perform well in both scenarios. 
\\ [1mm]
Since AUC is still influenced by the budget, we define a set of rules to set this hyperparameter upfront, so that we are not favoring a subset of algorithms by handcrafting a budget.
In this work, we choose the budget per dataset to be the first point at which one of 2 stopping conditions apply: (i) an algorithm (except Oracle) manages to reach 99\% of the full-dataset-performance (using the smallest query size) or (ii) the best algorithm (except oracle) did not improve the classifier's accuracy by at least 2\% in the last 20\% of iterations. \\ [1mm]
%\footnote{Since we only consider single-sample AL, increasing the budget comes at a high computational cost. Therefore we could not freely increase the budget above 2000.} 
As described in Sec. \ref{sec:restarts}, we will restart each experiment multiple times.
Each restart retains the train/test split (often given by the dataset itself), but creates a new validation split that is sampled (based on $s_\D$) from the entire dataset (not just the seed set $\LL_0$). \\ [1mm]
%
Apart from plotting standard performance curves and reporting their AUC values per dataset in App. \ref{app:all_results}, we primarily rely on ranks to aggregate the performance of an acquisition function across datasets.
For each dataset and query size, the AUC values of all acquisition functions are sorted and assigned a rank based on position, with the best rank being 1.
These ranks can safely be averages across datasets as they are no longer subjected to scaling differences of each dataset.
Additionally, we employ Critical Difference (CD) diagrams (like Fig. \ref{fig:ranks_by_domain}) for statistical testing.
CD diagrams use the Wilcoxon signed-rank test, which is a variant of the paired T-test, to find significant differences of ranks between acquisition functions.
For these diagrams, each combination of dataset, query size and run is considered a separate experiment, i.e. the results of \verb|Dataset1-QuerySize1-run5| of an acquisition function \verb|x| is only compared to the results of \verb|Dataset1-QuerySize1-run5| of acquisition function \verb|y|.
Due to the large number of restarts and the wide range of datasets and query sizes, we can provide very accurate significance tests.
For a detailed description of how every CD diagram is created, please refer to App. \ref{app:cd_diagrams}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Results on Synthetic Data}
Our results on Honeypot reveal principled shortcomings for the two best algorithms in BADGE and margin sampling.
Both are vulnerable to adverse samples or simply measurement noise, which highlights the need for further research in this area. \\
Finally, the fact that BADGE is able to perform well on Diverging Sine highlights the importance of embeddings for the clustering algorithms, as the so-called gradient embedding from BADGE seems to be able to encode uncertainty information, guiding the selection into the left hand regions of the dataset. 
We also show that embeddings are generally useful for this dataset, by providing results for "Coreset Raw" and "TypiClust Raw". \\ [1mm]
In conclusion, we strongly advocate to test newly proposed AL algorithms not only on a wide variety of real data domains, but also to pay close attention to the Honeypot and Diverging Sine datasets to reveal principled shortcomings of the algorithm in question.

\section{Conclusion}
TODO

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Description}
Given two spaces $\X, \Y$, $n=l+u$ data points with $l \in \mathbb{N}$ labeled examples $\mathcal{L} = \{(x_1, y_1),\ldots, (x_l,y_l)\}$, $u \in \mathbb{N}$ unlabeled examples $\mathcal{U} = \{x_{l+1},\ldots,x_{n}\}$, a model $\hat y: \X \to \Y$, a budget $\mathbb{N} \ni b \le u$ and an annotator $A: \mathcal{X} \to \mathcal{Y}$ that can label $x$. % (we consider only hard labels in the one-hot format). 
We call $x \in \mathcal{X}$, $y \in \mathcal{Y}$ predictors and labels respectively where $(x,y)$ are drawn from an unknown distribution $\rho$. 
Find an acquisition function $\Omega: \U^{(i)},\LL^{(i)} \mapsto x^{(i)} \in \U^{(i)}$ that iteratively selects the next unlabeled point $x^{(i)}$ for labeling
\begin{align*}
	\LL^{(i+1)} &\gets \LL^{(i)} \cup \{\left(x^{(i)}, A(x^{(i)})\right)\} \\
	\U^{(i+1)} &\gets \U^{(i)} \setminus \left\{x^{(i)}\right\} %, i \in 1{:}B
\end{align*}
with $\U^{(0)} = \operatorname{seed}(\U, s)$ and $\LL^{(0)} = \left(\U^{(0)}_i, A(\U^{(0)}_i)\right) \hspace{1mm} i \in [1, \ldots, s]$, where $\operatorname{seed}(\U, s)$ selects $s$ points per class for the initial labeled set. \\
So that the average expected loss $\ell: \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$ of a machine learning algorithm fitting $\hat y^{(i)}$ on the respective labeled set $\LL^{(i)}$ is minimal: 
$$\min \quad \frac{1}{B} \sum\limits_{i=0}^B \mathbb{E}_{(x,y) \sim \rho} \ell(y, \hat y^{(i)})$$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why we need 50 restarts}\label{sec:restarts}
To evaluate how many restarts are necessary to obtain conclusive results in an AL experiment, we computed 100 runs of our top-performing algorithm on one dataset.
Our best algorithm is margin sampling and we chose the Splice dataset for its average size and complexity. \\
This allows us firstly, to obtain a very strong estimation of the "true" average performance of margin sampling on this dataset and secondly, to draw subsets from this pool of 100 runs.
Setting the size of our draws to $\alpha$ and sampling uniformly, we can approximate a cross-validation process with $\alpha$ restarts.
Each of these draws can be interpreted as a \textbf{reported result in AL literature} where the authors employed $\alpha$ restarts.
%We provide evidence in Fig. \ref{fig:restarts} that previous works might not have evaluated their experiments with a sufficient number of restarts.
Figure \ref{fig:restarts} shows the "true" mean performance of margin sampling (green) in relation to random sampling (black) and the oracle performance (red).
We display 5 random draws of size $\alpha$ in blue.
We can observe that even for a relatively high number of restarts the variance between the samples is extremely high, resulting in some performance curves being worse that random and some being significantly better.
When setting $\alpha = 50$ we observe all samples to converge close to the true mean performance. 
\begin{figure}
        \hspace{-20mm}
	\includegraphics[width=1.24\linewidth]{img/ablation_restarts}
	\caption{Random draws from a pool of 100 runs for margin sampling on the Splice dataset with different numbers of repetitions ($\alpha=\{5,10,50\}$). Green curves are the mean performance of all 100 runs, while the samples are blue. Even with 5 or 10 repetitions, we can observe that single draws for margin sampling display below-random performance (black), while the true mean should be above random.}
	\label{fig:restarts}
\end{figure}
In addition to this motivating example, we carried out our main evaluation (Tab. \ref{tab:results}) multiple times by uniformly sampling 3 random from our 50 available runs and comparing the results.
We found significant differences in the performance of acquisition functions on individual datasets, as well as permutations in the final ranking.
This partly explains the ongoing difficulties in reproducing results for AL experiments and benchmarks.
This details can be found in App. \ref{app:rank_difference}.
For this benchmark we employ 50 restarts of every experiment.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation Protocol}\label{sec:evaluation}
Following \cite{zhou2021towards}, the quality of an AL algorithm is evaluated by an ``anytime protocol" that incorporates classification performance at every iteration, as opposed to evaluating final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{AUC}(\D_\test, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(\D_\test, \hat y^{(i)})
\end{equation}
The AUC incorporates performance in early stages (low budget) as well as capabilities to push the classifier in later stages (high budget).
AL algorithms have to perform well in both scenarios. 
\\ [1mm]
Since AUC is still influenced by the budget, we define a set of rules to set this hyperparameter upfront, so that we are not favoring a subset of algorithms by handcrafting a budget.
In this work, we choose the budget per dataset to be the first point at which one of 2 stopping conditions apply: (i) an algorithm (except Oracle) manages to reach 99\% of the full-dataset-performance (using the smallest query size) or (ii) the best algorithm (except oracle) did not improve the classifier's accuracy by at least 2\% in the last 20\% of iterations.
The first rule follows \cite{ji2023randomness}, while the second rule prevents excessive budgets for cases with diminishing returns in the budget.
The resulting budgets can be found in Tab. \ref{tab:batch_sizes}. \\ [1mm]
As described in Sec. \ref{sec:restarts}, we restart each experiment multiple times.
Each restart retains the train/test split (often given by the dataset itself), but creates a new validation split that is sampled (based on $s_\D$) from the entire dataset (not just the seed set $\LL^{(0)}$). \\ [1mm]
%
Apart from plotting standard performance curves and reporting their AUC values per dataset in App. \ref{app:all_results}, we primarily rely on ranks to aggregate the performance of an acquisition function across datasets.
For each dataset and query size, the AUC values of all acquisition functions are sorted and assigned a rank based on position, with the best rank being 1.
These ranks can safely be averages across datasets as they are no longer subjected to scaling differences of each dataset.
Additionally, we employ Critical Difference (CD) diagrams (like Fig. \ref{fig:ranks_by_domain}) for statistical testing.
CD diagrams use the Wilcoxon signed-rank test, which is a variant of the paired T-test, to find significant differences of ranks between acquisition functions.
For these diagrams, each combination of dataset, query size and run is considered a separate experiment, i.e. the results of \verb|Dataset1-QuerySize1-run5| of an acquisition function \verb|x| is only compared to the results of \verb|Dataset1-QuerySize1-run5| of acquisition function \verb|y|.
Due to the large number of restarts and the wide range of datasets and query sizes, we can provide very accurate significance tests.
For a detailed description of how every CD diagram is created, please refer to App. \ref{app:cd_diagrams}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Results on Synthetic Data}
Our results on Honeypot reveal principled shortcomings for the two best algorithms in BADGE and margin sampling.
Both are vulnerable to adverse samples or simply measurement noise, which highlights the need for further research in this area. \\
Finally, the fact that BADGE is able to perform well on Diverging Sine highlights the importance of embeddings for the clustering algorithms, as the so-called gradient embedding from BADGE seems to be able to encode uncertainty information, guiding the selection into the left hand regions of the dataset. 
We also show that embeddings are generally useful for this dataset, by providing results for "Coreset Raw" and "TypiClust Raw".

\section{Conclusion}
We strongly advocate to test newly proposed AL algorithms not only on a wide variety of real data domains, but also to pay close attention to the Honeypot and Diverging Sine datasets to reveal principled shortcomings of the algorithm in question.
Both tasks can be easily carried out by implementing the new acquisition function into our code base.
